---
title: "Constructing and Analysing the Ghana Sovereign Yield Curve, 2023–2026"
subtitle: "What the price of government debt says about Ghana's recovery from the 2022–23 debt crisis"
author: "Derrick Delali Atiase"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    number_sections: true
    fig_caption: true
    code_folding: show
    theme: flatly
    df_print: kable
always_allow_html: true
params:
  data_path: "Ghana_Yield_Curve_Dataset_2023_2026.xlsx"
  fit_frequency: "weekly"   # "weekly" (189 curves, fast) or "daily" (875 curves)
---

```{r setup, include=FALSE}
.fmt <- if (identical(knitr::opts_knit$get("rmarkdown.pandoc.to"), "html")) "html" else "gfm"
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.4, dpi = 110,
                      fig.align = "center", cache = TRUE, cache.lazy = FALSE,
                      cache.path = paste0("_cache/", .fmt, "/"),
                      fig.path = paste0("report_files/", .fmt, "/"))
library(readxl); library(dplyr); library(tidyr); library(ggplot2)
library(lubridate); library(purrr); library(stringr); library(scales)
library(knitr)
if (.fmt == "html") library(kableExtra)   # attaching it adds HTML dependencies

navy <- "#1B2A4A"; gold <- "#C8912A"; red <- "#B23A48"; grey <- "#8A8F98"
theme_set(theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(colour = "grey30", size = 9),
        plot.caption = element_text(colour = "grey40", size = 8, hjust = 0),
        panel.grid.minor = element_blank(),
        legend.position = "bottom"))

# tables render as styled HTML when knitting to HTML and as plain pipe tables
# when knitting to GitHub markdown, so the same source works for both outputs
tbl <- function(x, caption, digits = 2, ...) {
  if (.fmt == "html") {
    kbl(x, caption = caption, digits = digits, booktabs = TRUE,
        format.args = list(big.mark = ","), ...) |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                    full_width = FALSE, position = "center")
  } else {
    kable(x, caption = caption, digits = digits,
          format.args = list(big.mark = ","))
  }
}

wide <- function(k, col = 2, width = "30em") {
  if (.fmt == "html") column_spec(k, col, width = width) else k
}
SRC <- "Source: author's calculations on GFIM daily trading reports, Bank of Ghana and Ghana Statistical Service data."
```

# Why this project exists {#motivation}

In December 2022 Ghana defaulted on its domestic debt. The Domestic Debt
Exchange Programme (DDEP) that followed in February 2023 swapped most existing
cedi government bonds for a new set of longer, lower-coupon bonds. Inflation
peaked above 50%, the policy rate reached 30%, and the price of government
borrowing stopped behaving like a normal market.

A **yield curve** is simply the answer to one question asked at many horizons:
*what interest rate does the government have to pay to borrow for three months,
one year, five years, ten years?* Plotting those rates against their maturities
gives a single picture of the cost of money in an economy. It matters because:

- The **government** uses it to decide how much debt to issue and at what tenor.
- **Banks and pension funds** use it to value portfolios and measure interest
  rate risk; in Ghana, government securities are the largest asset class they
  hold.
- **Companies** price loans and corporate bonds off it.
- For **ordinary savers**, it is the benchmark behind Treasury bill returns and
  deposit rates.

Ghana has no published official zero-coupon curve. What exists is a daily file
of individual security quotes from the Ghana Fixed Income Market (GFIM) and
weekly Treasury bill auction results from the Bank of Ghana. Turning those into
a consistent curve, day by day, through a sovereign default and its recovery, is
the problem this project solves.

## Objectives {#objectives}

1. **Assemble** a clean daily dataset of every traded and quoted cedi
   government security from January 2023 to October 2026, joined to the policy
   rate, Treasury bill auction results and inflation.
2. **Establish the market conventions empirically** — the day-count basis for
   bills, the coupon frequency and day count for bonds, and whether published
   prices are clean or dirty — rather than assuming them.
3. **Fit a zero-coupon curve** for every trading day using bootstrapping,
   Nelson-Siegel and Svensson, and report how well each fits.
4. **Describe how the curve evolved** through the post-default recovery, and
   relate its level, slope and curvature to the policy rate and inflation.
5. **Apply the curve**: identify securities trading rich or cheap against it,
   and measure what a portfolio of government bonds loses or gains when rates
   move.

## What this report finds, in brief {#preview}

The curve starts in 2023 high and distorted, with short-dated borrowing priced
as dearly as long-dated borrowing, and ends in 2026 lower and upward sloping.
Short rates fall from roughly forty percent to the mid single digits. Real returns to savers flip from deeply negative to clearly
positive. A single "level" factor drives almost all day-to-day movement, which
means a bond portfolio's main risk is a general rise in rates rather than a
change in the curve's shape. And the quality of the market itself improves
sharply: quotes that scattered widely in 2023 line up tightly on a smooth curve
by 2026. Each of these is quantified below.

# Data {#data}

## Sources

```{r sources}
tibble(
  Source = c("GFIM daily trading reports", "GFIM / CSD security static data",
             "Bank of Ghana Treasury bill rates", "Bank of Ghana MPC decisions",
             "Ghana Statistical Service CPI"),
  Content = c("Daily price, yield, volume per security, 875 files",
              "ISIN, coupon, maturity, tenor",
              "Weekly primary auction discount and interest rates",
              "Policy rate decisions and effective dates",
              "Monthly headline year-on-year inflation"),
  Period = c("Jan 2023 – Oct 2026", "Current universe", "Aug 2013 – Sep 2026",
             "2002 – Sep 2026", "Jan 2023 – Aug 2026")
) |> tbl("Data sources used in the project.")
```

```{r load-data}
path <- params$data_path

quotes <- read_excel(path, sheet = "Curve_Inputs", guess_max = 20000) |>
  rename_with(~ tolower(gsub("[^a-z0-9]+", "_", tolower(.x)))) |>
  mutate(report_date = as.Date(report_date),
         maturity_date = as.Date(maturity_date),
         traded = as.logical(traded),
         use_for_fitting = as.logical(use_for_fitting))

panel <- read_excel(path, sheet = "Monthly_Panel") |>
  rename_with(~ tolower(gsub("[^a-z0-9]+", "_", tolower(.x)))) |>
  mutate(date = ceiling_date(as.Date(paste0(month, "-01")), "month") - 1)

cat(sprintf("%s quote rows | %s securities | %s trading days | %s to %s\n",
            format(nrow(quotes), big.mark = ","), n_distinct(quotes$isin),
            n_distinct(quotes$report_date), min(quotes$report_date),
            max(quotes$report_date)))
```

## What the market looks like

```{r coverage}
quotes |>
  group_by(Segment = segment) |>
  summarise(`Quote rows` = n(), Securities = n_distinct(isin),
            `Traded rows` = sum(traded),
            `Traded share` = percent(mean(traded), accuracy = 0.1),
            `Median yield (%)` = round(median(yield_pct, na.rm = TRUE), 2),
            .groups = "drop") |>
  arrange(desc(`Quote rows`)) |>
  tbl("Coverage and liquidity by market segment, January 2023 to October 2026. Segments are Treasury bills, post-DDEP bonds (ddep, gog_new), the pre-2023 combined sheet (gog) and pre-DDEP bonds (gog_old).")
```

Table \@ref(tab:coverage) carries two warnings for everything that follows.
First, **a quote is not a trade**: only about four rows in ten represent an
actual transaction, the rest being closing marks carried over from an earlier
session. Second, the **pre-DDEP bonds (`gog_old`) show a far higher median yield**
than post-DDEP bonds of similar maturity, on almost no trading. Section
\@ref(universe) shows those are stale marks rather than genuine bargains.

```{r liquidity-year}
quotes |>
  mutate(Year = as.character(year(report_date))) |>
  group_by(Year, Instrument = instrument) |>
  summarise(Quotes = n(), Traded = sum(traded),
            `Traded share` = percent(mean(traded), accuracy = 0.1),
            `Volume (GHS bn)` = round(sum(volume, na.rm = TRUE) / 1e9, 1),
            .groups = "drop") |>
  tbl("Quotes, trades and turnover by year and instrument type.")
```

Bills dominate turnover in every year, which is a feature of the market rather
than a quirk of the data: after the default, investors stayed short.

# Method {#method}

The curve cannot be fitted until we know how each instrument is quoted. Rather
than assume conventions, we test them against data the market publishes twice
over: both a price and a yield.

## Treasury bills: which day-count basis? {#bill-convention}

```{r bill-functions}
BILL_BASIS <- 364

bill_price <- function(yield_pct, days, basis = BILL_BASIS)
  100 / (1 + yield_pct / 100 * days / basis)

bill_yield <- function(price, days, basis = BILL_BASIS)
  (100 / price - 1) * basis / days * 100
```

```{r bill-convention}
bt <- quotes |>
  filter(instrument == "BILL", !is.na(closing_price), !is.na(closing_yield),
         ttm_days > 0, closing_price > 0)

tibble(`Day-count basis` = c("Actual/360", "Actual/364", "Actual/365")) |>
  mutate(`Median absolute error (pp)` = map_dbl(c(360, 364, 365), ~ median(abs(
    bill_yield(bt$closing_price, bt$ttm_days, .x) - bt$closing_yield)))) |>
  tbl(paste0("Testing the bill convention: median gap between the yield implied by the published price and the published yield itself, across ",
             format(nrow(bt), big.mark = ","), " quotes."), digits = 4)
```

Actual/364 reproduces GFIM's published bill yields **exactly**. The other two
bases do not. In plain terms, a Ghanaian Treasury bill is quoted as a simple
money-market yield on a 364-day year, and that single fact lets us recover
yields for the 2023 and 2024 reports, which published prices only.

## Government bonds: frequency, day count, clean or dirty? {#bond-convention}

```{r bond-functions}
add_months_safe <- function(d, n) d %m+% months(n)

coupon_dates <- function(settle, maturity, freq = 2) {
  step <- 12 / freq; out <- c(); d <- maturity
  while (d > settle) { out <- c(out, d); d <- add_months_safe(d, -step) }
  sort(as.Date(out, origin = "1970-01-01"))
}

yearfrac <- function(from, to, convention = "act/365") {
  switch(convention,
         "act/365" = as.numeric(to - from) / 365,
         "30/360"  = { d1 <- pmin(day(from), 30); d2 <- pmin(day(to), 30)
           ((year(to) - year(from)) * 360 + (month(to) - month(from)) * 30 +
              (d2 - d1)) / 360 },
         stop("unknown convention"))
}

accrued_interest <- function(settle, maturity, coupon, freq = 2,
                             convention = "act/365") {
  if (is.na(coupon) || coupon == 0) return(0)
  nxt <- coupon_dates(settle, maturity, freq)[1]
  prev <- add_months_safe(nxt, -12 / freq)
  period <- yearfrac(prev, nxt, convention)
  if (period <= 0) return(0)
  coupon / freq * yearfrac(prev, settle, convention) / period
}

bond_clean_price <- function(ytm, settle, maturity, coupon, freq = 2,
                             convention = "act/365") {
  d <- coupon_dates(settle, maturity, freq)
  if (!length(d)) return(NA_real_)
  t <- yearfrac(settle, d, convention) * freq
  cf <- rep(coupon / freq, length(d)); cf[length(cf)] <- cf[length(cf)] + 100
  sum(cf / (1 + ytm / 100 / freq)^t) -
    accrued_interest(settle, maturity, coupon, freq, convention)
}

bond_ytm <- function(clean, settle, maturity, coupon, freq = 2,
                     convention = "act/365") {
  f <- function(y) bond_clean_price(y, settle, maturity, coupon, freq,
                                    convention) - clean
  tryCatch(uniroot(f, c(-20, 300), tol = 1e-8)$root, error = function(e) NA_real_)
}
```

```{r bond-convention}
set.seed(1)
bs <- quotes |>
  filter(instrument == "BOND", use_for_fitting, !is.na(coupon_pct),
         ttm_years > 0.6, between(closing_price, 20, 160),
         between(yield_pct, 1, 80)) |>
  slice_sample(n = 800)

grid <- expand_grid(Frequency = c("Annual", "Semi-annual"),
                    `Day count` = c("act/365", "30/360")) |>
  mutate(freq = ifelse(Frequency == "Annual", 1, 2))

grid$`Median absolute price error` <- pmap_dbl(
  list(grid$freq, grid$`Day count`), function(f, conv) {
    p <- pmap_dbl(list(bs$yield_pct, bs$report_date, bs$maturity_date, bs$coupon_pct),
                  ~ bond_clean_price(..1, ..2, ..3, ..4, f, conv))
    median(abs(p - bs$closing_price), na.rm = TRUE)
  })

grid |> select(-freq) |>
  tbl("Testing the bond convention on 800 randomly sampled quotes: median gap between the price implied by the published yield and the published price, per 100 face value.", digits = 4)
```

```{r clean-dirty}
model_clean <- pmap_dbl(list(bs$yield_pct, bs$report_date, bs$maturity_date, bs$coupon_pct),
                        ~ bond_clean_price(..1, ..2, ..3, ..4))
ai <- pmap_dbl(list(bs$report_date, bs$maturity_date, bs$coupon_pct),
               ~ accrued_interest(..1, ..2, ..3))
clean_err <- model_clean - bs$closing_price

tibble(Assumption = c("Published price is clean (all quotes)",
                      "Published price is clean (traded quotes only)",
                      "Published price is dirty (all quotes)"),
       `Median absolute error` = c(median(abs(clean_err), na.rm = TRUE),
                                   median(abs(clean_err[bs$traded]), na.rm = TRUE),
                                   median(abs(model_clean + ai - bs$closing_price), na.rm = TRUE))) |>
  tbl("Is the published bond price clean or dirty? Median absolute pricing error per 100 face under each assumption.", digits = 4)
```

Tables \@ref(tab:bond-convention) and \@ref(tab:clean-dirty) settle the bond
convention: **semi-annual coupons, actual/365, quoted clean**. The pricing error
is about 0.11 per 100 face across all quotes but only
`r round(median(abs(clean_err[bs$traded]), na.rm = TRUE), 3)` on quotes that
actually traded. That difference is informative in itself: the residual error on
untraded lines is staleness, not a flaw in the pricing model, and it justifies
weighting traded quotes more heavily when fitting.

## Fitting the curve {#fitting}

Each day's securities are turned into cash flows: a bill is one payment at
maturity, a bond is a stream of semi-annual coupons plus principal. A candidate
curve discounts those cash flows; the fitted curve is the one whose prices come
closest to market prices, with three safeguards:

- quotes more than three robust deviations from the local median yield are
  dropped, so one bad print cannot bend the curve;
- each quote is weighted by the inverse of its maturity, so a price error at ten
  years is not allowed to swamp an equal-sized error at three months;
- quotes that did not trade that day are weighted at 35%.

```{r fit-machinery}
UNIVERSE <- c("bills", "ddep", "gog_new", "gog")
UNTRADED_W <- 0.35
GRID <- c(0.25, 0.5, 0.75, 1, 2, 3, 4, 5, 7, 10, 12, 15)

build_day <- function(rows, settle) {
  rows <- rows |> filter(segment %in% UNIVERSE, use_for_fitting,
                         !is.na(closing_price), closing_price > 0,
                         ttm_years > 0) |> arrange(ttm_years)
  if (nrow(rows) < 10) return(NULL)
  k <- min(9, 2 * floor(nrow(rows) / 2) - 1)
  resid <- abs(rows$yield_pct - stats::runmed(rows$yield_pct, k = k))
  mad <- median(resid, na.rm = TRUE)
  rows <- rows[resid <= 3 * 1.4826 * mad + 0.75, ]
  if (nrow(rows) < 8) return(NULL)

  flows <- map_dfr(seq_len(nrow(rows)), function(i) {
    r <- rows[i, ]
    if (r$instrument == "BILL") return(tibble(id = i, t = r$ttm_years, cf = 100))
    d <- coupon_dates(settle, r$maturity_date)
    if (!length(d)) return(NULL)
    cf <- rep(r$coupon_pct / 2, length(d)); cf[length(cf)] <- cf[length(cf)] + 100
    tibble(id = i, t = yearfrac(settle, d), cf = cf)
  })
  price <- map_dbl(seq_len(nrow(rows)), function(i) {
    r <- rows[i, ]
    if (r$instrument == "BILL") r$closing_price
    else r$closing_price + accrued_interest(settle, r$maturity_date, r$coupon_pct)
  })
  list(rows = rows, flows = flows, price = price,
       weight = (1 / pmax(rows$ttm_years, 0.1)) *
         ifelse(rows$traded, 1, UNTRADED_W))
}

ns_zero <- function(t, b0, b1, b2, tau) {
  x <- pmax(t, 1e-6) / tau; decay <- (1 - exp(-x)) / x
  b0 + b1 * decay + b2 * (decay - exp(-x))
}
sv_zero <- function(t, b0, b1, b2, b3, tau1, tau2) {
  x1 <- pmax(t, 1e-6) / tau1; x2 <- pmax(t, 1e-6) / tau2
  d1 <- (1 - exp(-x1)) / x1; d2 <- (1 - exp(-x2)) / x2
  b0 + b1 * d1 + b2 * (d1 - exp(-x1)) + b3 * (d2 - exp(-x2))
}
model_prices <- function(day, zfn) {
  z <- zfn(day$flows$t) / 100
  as.numeric(tapply(day$flows$cf * exp(-z * day$flows$t), day$flows$id, sum))
}
huber <- function(r, k = 0.3) ifelse(abs(r) <= k, 0.5 * r^2, k * (abs(r) - 0.5 * k))

fit_ns <- function(day) {
  y <- day$rows$yield_pct
  short <- median(y[day$rows$ttm_years < 1], na.rm = TRUE)
  long  <- median(y[day$rows$ttm_years > 5], na.rm = TRUE)
  if (is.na(short)) short <- median(y, na.rm = TRUE)
  if (is.na(long))  long  <- short
  obj <- function(p) sum(huber((model_prices(
    day, function(t) ns_zero(t, p[1], p[2], p[3], p[4])) - day$price) * day$weight))
  optim(c(long, short - long, 0, 1.5), obj, method = "L-BFGS-B",
        lower = c(-50, -100, -100, 0.05), upper = c(200, 100, 100, 30),
        control = list(maxit = 500))$par
}

fit_sv <- function(day, start) {
  obj <- function(p) sum(huber((model_prices(
    day, function(t) sv_zero(t, p[1], p[2], p[3], p[4], p[5], p[6])) - day$price) * day$weight))
  optim(c(start[1:3], 0, start[4], max(start[4] * 3, 5)), obj, method = "L-BFGS-B",
        lower = c(-50, -100, -100, -100, 0.05, 0.05),
        upper = c(200, 100, 100, 100, 30, 30), control = list(maxit = 800))$par
}

bootstrap_curve <- function(day) {
  bills <- which(day$rows$instrument == "BILL")
  kt <- day$rows$ttm_years[bills]
  kz <- -log(day$price[bills] / 100) / kt * 100
  ord <- order(kt); kt <- kt[ord]; kz <- kz[ord]
  bonds <- which(day$rows$instrument == "BOND")
  bonds <- bonds[order(day$rows$ttm_years[bonds])]
  for (i in bonds) {
    tmax <- day$rows$ttm_years[i]
    if (tmax <= max(kt)) next
    f <- function(zl) {
      kt2 <- c(kt, tmax); kz2 <- c(kz, zl)
      fl <- day$flows[day$flows$id == i, ]
      z <- approx(kt2, kz2, fl$t, rule = 2)$y / 100
      sum(fl$cf * exp(-z * fl$t)) - day$price[i]
    }
    zl <- tryCatch(uniroot(f, c(-20, 200), tol = 1e-8)$root, error = function(e) NA)
    if (!is.na(zl)) { kt <- c(kt, tmax); kz <- c(kz, zl) }
  }
  list(t = kt, z = kz, fn = function(tt) approx(kt, kz, tt, rule = 2)$y)
}

fit_quality <- function(day, zfn) {
  e <- model_prices(day, zfn) - day$price
  c(rmse_price = sqrt(mean(e^2)),
    rmse_yield_bp = sqrt(mean((e * day$weight)^2)) * 100)
}
```

## A single day, in detail {#single-day}

```{r single-day}
last_date <- max(quotes$report_date)
day <- build_day(filter(quotes, report_date == last_date), last_date)
p_ns <- fit_ns(day)
p_sv <- fit_sv(day, p_ns)
bs_curve <- bootstrap_curve(day)

tibble(Method = c("Nelson-Siegel", "Svensson", "Bootstrap"),
       `RMSE (bp)` = c(fit_quality(day, function(t) ns_zero(t, p_ns[1], p_ns[2], p_ns[3], p_ns[4]))[["rmse_yield_bp"]],
                       fit_quality(day, function(t) sv_zero(t, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6]))[["rmse_yield_bp"]],
                       fit_quality(day, bs_curve$fn)[["rmse_yield_bp"]]),
       Parameters = c(4, 6, length(bs_curve$t)),
       `1-year (%)` = c(ns_zero(1, p_ns[1], p_ns[2], p_ns[3], p_ns[4]),
                        sv_zero(1, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6]),
                        bs_curve$fn(1)),
       `5-year (%)` = c(ns_zero(5, p_ns[1], p_ns[2], p_ns[3], p_ns[4]),
                        sv_zero(5, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6]),
                        bs_curve$fn(5)),
       `10-year (%)` = c(ns_zero(10, p_ns[1], p_ns[2], p_ns[3], p_ns[4]),
                         sv_zero(10, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6]),
                         bs_curve$fn(10))) |>
  tbl(paste("Three fitting methods compared on", format(last_date, "%d %B %Y"),
            "- fit error, number of parameters and the fitted zero rate at three maturities."))
```

```{r fig-methods, fig.cap="The three fitting methods compared on the last trading day. Top panel: the fitted zero curves against the observed quotes (shown as yields, which sit slightly above continuously compounded zero rates at the same maturity). Bottom panel: each method's pricing error per security, converted to basis points of yield. The bootstrap passes through the bill points by construction; the two parametric curves trade exactness for smoothness.", fig.height=6.2}
library(patchwork)

tt <- seq(0.05, max(day$rows$ttm_years), length.out = 300)
curves_df <- bind_rows(
  tibble(t = tt, z = ns_zero(tt, p_ns[1], p_ns[2], p_ns[3], p_ns[4]), Method = "Nelson-Siegel"),
  tibble(t = tt, z = sv_zero(tt, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6]), Method = "Svensson"),
  tibble(t = tt, z = bs_curve$fn(tt), Method = "Bootstrap"))

obs_df <- day$rows |> transmute(t = ttm_years, y = yield_pct,
                                Traded = ifelse(traded, "Traded", "Not traded"))

method_cols <- c(`Nelson-Siegel` = navy, Svensson = gold, Bootstrap = red)

p_curves <- ggplot() +
  geom_point(data = obs_df, aes(t, y, shape = Traded), colour = grey,
             size = 1.6, alpha = .7) +
  geom_line(data = curves_df, aes(t, z, colour = Method, linetype = Method),
            linewidth = .9) +
  scale_colour_manual(values = method_cols) +
  scale_linetype_manual(values = c(`Nelson-Siegel` = 1, Svensson = 2, Bootstrap = 3)) +
  labs(title = "Three fits to the same day's quotes",
       subtitle = format(last_date, "%d %B %Y"),
       x = NULL, y = "Zero rate / yield (%)", shape = NULL)

resid_df <- bind_rows(
  tibble(t = day$rows$ttm_years, Method = "Nelson-Siegel",
         bp = (model_prices(day, function(x) ns_zero(x, p_ns[1], p_ns[2], p_ns[3], p_ns[4])) -
                 day$price) * day$weight * 100),
  tibble(t = day$rows$ttm_years, Method = "Svensson",
         bp = (model_prices(day, function(x) sv_zero(x, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6])) -
                 day$price) * day$weight * 100),
  tibble(t = day$rows$ttm_years, Method = "Bootstrap",
         bp = (model_prices(day, bs_curve$fn) - day$price) * day$weight * 100))

p_resid <- ggplot(resid_df, aes(t, bp, colour = Method)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point(size = 1.5, alpha = .8) +
  facet_wrap(~ Method, nrow = 1) +
  scale_colour_manual(values = method_cols, guide = "none") +
  labs(title = "Pricing errors by maturity",
       x = "Years to maturity", y = "Error (basis points)", caption = SRC)

p_curves / p_resid + plot_layout(heights = c(1.15, 1))
```

Figure \@ref(fig:fig-methods) shows what Table \@ref(tab:single-day) only
summarises. All three methods agree closely out to about five years, where the
bills and shorter bonds crowd together, and separate at the long end where the
data thins out. The residual panel explains the ranking: the bootstrap drives
errors on the bill points to zero because it is built to pass through them, but
it inherits every kink in the data and has as many parameters as instruments.
Svensson adds two parameters to Nelson-Siegel and buys very little here.
**Nelson-Siegel is used for the rest of the report**: four interpretable
parameters, a smooth curve, and fit error close to the alternatives.

```{r fig-methods-long, fig.cap="The same three fits extended to twenty years, with the actual market quotes shown as grey points. The shaded region marks maturities where no security trades, so the curves there are extrapolation. Nelson-Siegel and Svensson decay towards their long-run parameters while the bootstrap simply holds its last observed level flat.", fig.height=3.8}
tt2 <- seq(0.05, 20, length.out = 300)
maxt <- max(day$rows$ttm_years)

bind_rows(
  tibble(t = tt2, z = ns_zero(tt2, p_ns[1], p_ns[2], p_ns[3], p_ns[4]), Method = "Nelson-Siegel"),
  tibble(t = tt2, z = sv_zero(tt2, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6]), Method = "Svensson"),
  tibble(t = tt2, z = bs_curve$fn(tt2), Method = "Bootstrap")) |>
  ggplot(aes(t, z, colour = Method, linetype = Method)) +
  annotate("rect", xmin = maxt, xmax = 20, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .5) +
  geom_point(data = transmute(day$rows, t = ttm_years, z = yield_pct),
             aes(t, z), inherit.aes = FALSE, colour = grey, size = 1.4,
             alpha = .7) +
  annotate("text", x = (maxt + 20) / 2, y = Inf, vjust = 1.6,
           label = "no traded securities", size = 3, colour = "grey35") +
  geom_vline(xintercept = maxt, colour = "grey50", linetype = 3) +
  geom_line(linewidth = .9) +
  scale_colour_manual(values = method_cols) +
  scale_linetype_manual(values = c(`Nelson-Siegel` = 1, Svensson = 2, Bootstrap = 3)) +
  labs(title = "Where the methods disagree: the extrapolated long end",
       subtitle = paste("Longest traded security matures in", round(maxt, 1), "years"),
       x = "Years to maturity", y = "Zero rate (%)", caption = SRC)
```

```{r fig-methods-facet, fig.cap="Actual quotes against each fitted curve, one panel per method. Grey points are the market quotes that entered the fit; the coloured line is that method's curve. Plotting them separately makes it easier to see where each method tracks the data and where it smooths through it.", fig.height=3.6}
obs_rep <- map_dfr(names(method_cols), ~ mutate(obs_df, Method = .x)) |>
  mutate(Method = factor(Method, levels = names(method_cols)))

curves_df |> mutate(Method = factor(Method, levels = names(method_cols))) |>
  ggplot(aes(t, z)) +
  geom_point(data = obs_rep, aes(t, y), colour = grey, size = 1.2, alpha = .65) +
  geom_line(aes(colour = Method), linewidth = .9) +
  facet_wrap(~ Method, nrow = 1) +
  scale_colour_manual(values = method_cols, guide = "none") +
  labs(title = "Actual quotes and fitted curve, by method",
       subtitle = format(last_date, "%d %B %Y"),
       x = "Years to maturity", y = "Zero rate / yield (%)", caption = SRC)
```

```{r fig-actual-fitted, fig.cap="Actual against fitted prices for every security in the fit. Each point is one security: the horizontal axis is the price the market quoted, the vertical axis the price the fitted curve implies. The dashed line is perfect agreement, so distance from it is the pricing error. Panel headings give the root mean squared error in cedis per 100 face.", fig.height=3.6}
fitted_prices <- bind_rows(
  tibble(Method = "Nelson-Siegel", actual = day$price,
         fitted = model_prices(day, function(x) ns_zero(x, p_ns[1], p_ns[2], p_ns[3], p_ns[4]))),
  tibble(Method = "Svensson", actual = day$price,
         fitted = model_prices(day, function(x) sv_zero(x, p_sv[1], p_sv[2], p_sv[3], p_sv[4], p_sv[5], p_sv[6]))),
  tibble(Method = "Bootstrap", actual = day$price,
         fitted = model_prices(day, bs_curve$fn))) |>
  mutate(Method = factor(Method, levels = names(method_cols)))

lab <- fitted_prices |> group_by(Method) |>
  summarise(rmse = sqrt(mean((fitted - actual)^2)), .groups = "drop") |>
  mutate(facet = sprintf("%s (RMSE %.2f)", Method, rmse))

fitted_prices |> left_join(lab, by = "Method") |>
  mutate(facet = factor(facet, levels = lab$facet)) |>
  ggplot(aes(actual, fitted, colour = Method)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(size = 1.5, alpha = .8) +
  facet_wrap(~ facet, nrow = 1) +
  scale_colour_manual(values = method_cols, guide = "none") +
  labs(title = "Actual against fitted prices",
       subtitle = paste("All securities in the fit,", format(last_date, "%d %B %Y")),
       x = "Market price (per 100 face)", y = "Fitted price (per 100 face)",
       caption = SRC)
```

Figures \@ref(fig:fig-methods-facet) and \@ref(fig:fig-actual-fitted) show the
same comparison two ways: against maturity, and actual against fitted price. In
the price scatter the bootstrap sits almost exactly on the 45-degree line, as it
must, while the two parametric curves show a visible spread. That spread is the
price of smoothness, and it is small enough in cedi terms to be worth paying.

The second panel is a caution for anyone quoting a fifteen-year Ghanaian rate.
Past about `r round(max(day$rows$ttm_years), 1)` years the three methods diverge
because nothing trades there, so the number depends on the model rather than the
market. Results in this report are quoted to ten years for that reason.

```{r fig-single-day, fig.cap="Observed market quotes and the fitted Nelson-Siegel curve on the last trading day of the sample. Each point is one security: circles are Treasury bills, triangles are bonds. Dark points traded that day; grey points are carried-over marks. The curve tracks the traded cluster rather than the stale quotes.", fig.height=4}
obs <- day$rows |> transmute(ttm_years, yield_pct,
                             Traded = ifelse(traded, "Traded", "Not traded"),
                             Instrument = instrument)
line <- tibble(t = seq(0.05, max(obs$ttm_years), length.out = 300)) |>
  mutate(z = ns_zero(t, p_ns[1], p_ns[2], p_ns[3], p_ns[4]))

ggplot() +
  geom_point(data = obs, aes(ttm_years, yield_pct, colour = Traded,
                             shape = Instrument), size = 2, alpha = .85) +
  geom_line(data = line, aes(t, z), colour = navy, linewidth = 1) +
  scale_colour_manual(values = c(Traded = navy, `Not traded` = grey)) +
  labs(title = paste("Fitted zero curve,", format(last_date, "%d %B %Y")),
       subtitle = "Nelson-Siegel curve against individual security quotes",
       x = "Years to maturity", y = "Yield / zero rate (%)", caption = SRC)
```

## Why pre-DDEP bonds are excluded {#universe}

```{r universe}
quotes |>
  filter(report_date == last_date,
         segment %in% c("bills", "ddep", "gog_new", "gog_old")) |>
  group_by(Segment = segment) |>
  summarise(Securities = n(), Traded = sum(traded),
            `Median yield (%)` = round(median(yield_pct, na.rm = TRUE), 2),
            `Longest maturity (yrs)` = round(max(ttm_years), 1), .groups = "drop") |>
  tbl(paste("Why pre-DDEP bonds are left out: segment comparison on",
            format(last_date, "%d %B %Y"), "- old bonds show much higher yields on almost no trading."))
```

A bond that has not traded for months but is still quoted at a stale price looks
cheap when it is merely forgotten. Including the `gog_old` bonds pulls the long
end of the fitted curve up by several percentage points without adding
information. They are excluded; reinstating them is a one-line change to
`UNIVERSE` for a robustness check.

# Results {#results}

## Fitting every trading day

```{r fit-all}
all_dates <- sort(unique(quotes$report_date))
fit_dates <- if (params$fit_frequency == "daily") all_dates else {
  tibble(d = all_dates) |> mutate(wk = floor_date(d, "week")) |>
    group_by(wk) |> slice_max(d, n = 1) |> pull(d)
}

by_date <- quotes |> filter(segment %in% UNIVERSE, use_for_fitting) |>
  group_split(report_date)
names(by_date) <- map_chr(by_date, ~ as.character(.x$report_date[1]))

fit_one <- function(key) {
  d <- as.Date(key); rows <- by_date[[key]]
  if (is.null(rows)) return(NULL)
  dd <- build_day(rows, d); if (is.null(dd)) return(NULL)
  p <- fit_ns(dd)
  q <- fit_quality(dd, function(t) ns_zero(t, p[1], p[2], p[3], p[4]))
  z <- as.list(ns_zero(GRID, p[1], p[2], p[3], p[4]))
  names(z) <- paste0("z", GRID, "y")
  bind_cols(tibble(report_date = d, n = nrow(dd$rows),
                   n_traded = sum(dd$rows$traded),
                   max_ttm = max(dd$rows$ttm_years),
                   beta0 = p[1], beta1 = p[2], beta2 = p[3], tau = p[4],
                   rmse_bp = q[["rmse_yield_bp"]]), as_tibble(z))
}

fits <- map_dfr(as.character(fit_dates), fit_one) |>
  mutate(level = beta0, slope = -beta1, curvature = beta2)

cat(sprintf("%d curves fitted (%s), %s to %s\n", nrow(fits),
            params$fit_frequency, min(fits$report_date), max(fits$report_date)))
```

## Finding 1: rates collapsed and the curve regained a normal slope {#finding-shape}

```{r fig-snapshots, fig.cap="The Ghana sovereign zero curve on one date from each year of the sample. The vertical shift between the lines is the fall in the general level of rates; the change in tilt is the return of a normal upward slope. Exact values are in Table 4.2."}
snaps <- c(min(fits$report_date),
           fits$report_date[which.min(abs(fits$report_date - as.Date("2024-06-30")))],
           fits$report_date[which.min(abs(fits$report_date - as.Date("2025-06-30")))],
           max(fits$report_date))

fits |> filter(report_date %in% snaps) |>
  select(report_date, starts_with("z")) |>
  pivot_longer(-report_date, names_to = "mat", values_to = "zero") |>
  mutate(maturity = as.numeric(str_remove_all(mat, "[zy]")),
         Date = factor(format(report_date, "%d %b %Y"),
                       levels = format(sort(snaps), "%d %b %Y"))) |>
  ggplot(aes(maturity, zero, colour = Date)) +
  geom_line(linewidth = 1) + geom_point(size = 1.8) +
  scale_colour_manual(values = c(red, gold, grey, navy)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(title = "Lower rates and a normal slope",
       subtitle = "Fitted zero-coupon curves, four dates",
       x = "Years to maturity", y = "Zero rate", colour = "Curve date",
       caption = SRC)
```

```{r tab-snapshots}
fits |> filter(report_date %in% snaps) |>
  transmute(Date = format(report_date, "%d %b %Y"),
            `3m` = z0.25y, `1y` = z1y, `2y` = z2y, `5y` = z5y,
            `10y` = z10y, `Slope 10y-1y (pp)` = z10y - z1y) |>
  tbl("Fitted zero rates (%) at selected maturities on the four snapshot dates.")
```

```{r snapshot-numbers, echo=FALSE}
snapvals <- fits |> filter(report_date %in% snaps) |> arrange(report_date) |>
  transmute(date = report_date, m3 = z0.25y, y1 = z1y, y10 = z10y,
            slope = z10y - z1y)
say <- function(i) with(snapvals[i, ], sprintf(
  "%s: %.1f%% at three months, %.1f%% at one year and %.1f%% at ten years (slope %+.1f points)",
  format(date, "%b %Y"), m3, y1, y10, slope))
```

Figure \@ref(fig:fig-snapshots) and Table \@ref(tab:tab-snapshots) are the core
result, and two things move at once: the **level** of rates and the **tilt** of
the curve.

- `r say(1)`. Borrowing was expensive at every horizon, and the very short end
  was the most distorted part of the curve.
- `r say(2)` — short rates still far above long rates, which is what a market
  looks like when near-term risk, not long-term growth, is the worry.
- `r say(3)`.
- `r say(4)`. Short rates have collapsed and the curve now rises with maturity
  over most of its length.

Two separate things are happening, and Table \@ref(tab:tab-slope) separates
them. The **level** falls in every year without reversal. The **slope** is
negative on average in 2023 and 2024, meaning it cost the government more to
borrow for one year than for ten, and turns positive in 2025 and 2026. Day to
day the 2023 slope is erratic, which is itself a symptom: when a market is
pricing default risk rather than the path of interest rates, the ordinary
relationship between maturity and yield breaks down.

```{r tab-slope}
fits |> mutate(Year = as.character(year(report_date))) |>
  group_by(Year) |>
  summarise(`3-month (%)` = mean(z0.25y), `1-year (%)` = mean(z1y),
            `10-year (%)` = mean(z10y),
            `Slope 10y-1y (pp)` = mean(z10y - z1y), .groups = "drop") |>
  tbl("Annual averages of the fitted curve. The level falls in every year; the slope is negative in 2023 and 2024 and positive thereafter.")
```

## Finding 2: the market itself became more orderly {#finding-quality}

```{r fig-fit-quality, fig.cap="Curve fit error over time. The line is the root mean squared error of each day's fit, expressed in basis points of yield. Because the same model is used throughout, the fall measures how much more consistently securities are priced relative to one another, not an improvement in the model."}
ggplot(fits, aes(report_date, rmse_bp)) +
  geom_line(colour = grey, linewidth = .4) +
  geom_smooth(se = FALSE, colour = gold, linewidth = 1.1, span = .25) +
  labs(title = "Quotes line up more tightly as the market normalises",
       subtitle = "Root mean squared fit error of the daily Nelson-Siegel curve",
       x = NULL, y = "Fit error (basis points)", caption = SRC)
```

```{r tab-fit-quality}
fits |> mutate(Year = as.character(year(report_date))) |>
  group_by(Year) |>
  summarise(Curves = n(), `Median fit error (bp)` = round(median(rmse_bp)),
            `Median securities` = median(n), `Median traded` = median(n_traded),
            .groups = "drop") |>
  tbl("Fit quality by year. Fit error falls by roughly a factor of six between 2023 and 2026 while the number of securities quoted rises.")
```

This is a result, not a technical footnote. A large fit error means securities
of similar maturity are quoted at very different yields, which happens when few
people are trading and marks are guesses. **The sixfold fall in fit error is a
measure of market repair.** It also means the 2023 curves should be read as
indicative, and any conclusion drawn from them stated with that caveat.

## Finding 3: level, slope and curvature {#finding-factors}

```{r fig-factors, fig.cap="The three Nelson-Siegel factors through time. Level is the long-run rate the curve tends to; slope is the gap between short and long rates (positive means short rates are below long rates); curvature captures the hump in the middle of the curve."}
fits |> select(report_date, Level = level, Slope = slope, Curvature = curvature) |>
  pivot_longer(-report_date, names_to = "Factor") |>
  ggplot(aes(report_date, value, colour = Factor)) +
  geom_hline(yintercept = 0, colour = "grey75", linewidth = .4) +
  geom_line(linewidth = .7) +
  scale_colour_manual(values = c(Level = navy, Slope = gold, Curvature = red)) +
  labs(title = "Nelson-Siegel factors",
       subtitle = "Fitted parameters of the daily curve",
       x = NULL, y = "Percentage points", caption = SRC)
```

```{r fig-pca, fig.cap="Principal component analysis of weekly changes in the fitted curve. PC1 moves all maturities in the same direction (a level shift), PC2 moves short and long rates in opposite directions (a slope change) and PC3 moves the middle against the ends (curvature)."}
z_weekly <- fits |> select(report_date, starts_with("z")) |> arrange(report_date)
dz <- z_weekly |> select(-report_date) |> as.matrix() |> diff()
pca <- prcomp(dz, center = TRUE, scale. = FALSE)
explained <- pca$sdev^2 / sum(pca$sdev^2)

as_tibble(pca$rotation[, 1:3], rownames = "mat") |>
  mutate(maturity = as.numeric(str_remove_all(mat, "[zy]"))) |>
  pivot_longer(PC1:PC3, names_to = "Component") |>
  ggplot(aes(maturity, value, colour = Component)) +
  geom_hline(yintercept = 0, colour = "grey75") +
  geom_line(linewidth = .9) + geom_point(size = 1.5) +
  scale_colour_manual(values = c(PC1 = navy, PC2 = gold, PC3 = red),
                      labels = c(paste0("PC1 level (", percent(explained[1], .1), ")"),
                                 paste0("PC2 slope (", percent(explained[2], .1), ")"),
                                 paste0("PC3 curvature (", percent(explained[3], .1), ")"))) +
  labs(title = "Three factors explain almost all curve movement",
       subtitle = "Loadings from PCA of weekly zero-rate changes",
       x = "Years to maturity", y = "Loading", caption = SRC)
```

```{r tab-pca}
tibble(Component = paste0("PC", 1:4),
       `Share of variance` = percent(explained[1:4], accuracy = 0.1),
       Interpretation = c("Level: all rates move together",
                          "Slope: short and long rates diverge",
                          "Curvature: the middle moves against the ends",
                          "Residual")) |>
  tbl("Share of weekly curve variation explained by each principal component.")
```

Three factors explain
`r percent(sum(explained[1:3]), accuracy = 0.1)` of weekly curve movement, and
the level factor alone explains `r percent(explained[1], accuracy = 0.1)`. That
level share is higher than the 60–70% typically reported for developed markets.
The practical reading: in Ghana, rates mostly move *together*, driven by one
macro force, so a bond portfolio's dominant risk is a general rise in rates
rather than a change in the curve's shape. Section \@ref(risk) puts a cedi
figure on that.

## Finding 4: the curve, the policy rate and inflation {#finding-macro}

```{r fig-macro, fig.cap="Fitted zero rates against the Bank of Ghana policy rate and headline inflation. Market rates fell ahead of the policy rate through 2025; the three-month rate only settles below the policy rate late in the sample."}
monthly <- fits |>
  mutate(date = ceiling_date(report_date, "month") - 1) |>
  group_by(date) |>
  summarise(`3-month zero` = mean(z0.25y), `1-year zero` = mean(z1y),
            `10-year zero` = mean(z10y), .groups = "drop") |>
  left_join(select(panel, date, `Policy rate` = policy_rate_pct,
                   `Headline inflation` = headline_inflation_pct), by = "date")

monthly |> pivot_longer(-date, names_to = "Series") |> filter(!is.na(value)) |>
  ggplot(aes(date, value, colour = Series, linetype = Series)) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(`3-month zero` = gold, `1-year zero` = navy,
                                 `10-year zero` = grey, `Policy rate` = red,
                                 `Headline inflation` = "black")) +
  scale_linetype_manual(values = c(`3-month zero` = 1, `1-year zero` = 1,
                                   `10-year zero` = 1, `Policy rate` = 2,
                                   `Headline inflation` = 3)) +
  guides(colour = guide_legend(nrow = 2)) +
  labs(title = "Market rates, policy rate and inflation",
       subtitle = "Monthly averages of the fitted curve against macro series",
       x = NULL, y = "Percent", caption = SRC)
```

```{r tab-real-rates}
real_tbl <- monthly |>
  mutate(real = `1-year zero` - `Headline inflation`) |>
  filter(!is.na(real))

bind_rows(head(real_tbl, 3), tail(real_tbl, 3)) |>
  transmute(Month = format(date, "%b %Y"),
            `1-year rate (%)` = round(`1-year zero`, 2),
            `Inflation (%)` = `Headline inflation`,
            `Real rate (pp)` = round(real, 2)) |>
  tbl("The one-year rate against inflation: first and last three months with inflation data. A negative real rate means a saver loses purchasing power even after interest.")
```

Table \@ref(tab:tab-real-rates) carries the clearest message for a
non-specialist. In early 2023 a saver lending to the government for a year
earned far less than prices rose, so money in Treasury bills **lost** real value.
By 2026 the same saver earns several points above inflation.

## Finding 5: what the curve expects next {#finding-forwards}

```{r fig-forwards, fig.cap="Spot and forward rates. The 1-year-in-1-year forward is the rate the market implies for borrowing one year, starting a year from now; the 5y5y forward is the ten-year horizon equivalent. When forwards sit above spot rates, the market expects rates to rise."}
fwd <- function(z1, t1, z2, t2) (z2 * t2 - z1 * t1) / (t2 - t1)

fits |>
  transmute(report_date,
            `1-year spot` = z1y,
            `1y in 1y forward` = fwd(z1y, 1, z2y, 2),
            `5y in 5y forward` = fwd(z5y, 5, z10y, 10)) |>
  pivot_longer(-report_date, names_to = "Series") |>
  ggplot(aes(report_date, value, colour = Series)) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(`1-year spot` = navy,
                                 `1y in 1y forward` = gold,
                                 `5y in 5y forward` = grey)) +
  labs(title = "Spot and implied forward rates",
       subtitle = "What today's curve implies about future borrowing costs",
       x = NULL, y = "Percent", caption = SRC)
```

# Applications {#applications}

## Which securities look rich or cheap? {#rich-cheap}

```{r rich-cheap}
zfn <- function(t) ns_zero(t, p_ns[1], p_ns[2], p_ns[3], p_ns[4])
mp <- model_prices(day, zfn)

model_yield <- vapply(seq_len(nrow(day$rows)), function(i) {
  r <- day$rows[i, ]
  if (r$instrument == "BILL") bill_yield(mp[i], max(round(r$ttm_years * 365), 1))
  else bond_ytm(mp[i] - accrued_interest(last_date, r$maturity_date, r$coupon_pct),
                last_date, r$maturity_date, r$coupon_pct)
}, numeric(1))

rc <- day$rows |>
  mutate(model_price = mp, market_price = day$price, model_yield = model_yield,
         spread_bp = round((yield_pct - model_yield) * 100, 1),
         signal = case_when(spread_bp > 50 ~ "Cheap", spread_bp < -50 ~ "Rich",
                            TRUE ~ "Fair")) |>
  select(isin, instrument, security, maturity_date, ttm_years, coupon_pct,
         yield_pct, model_yield, spread_bp, traded, volume, signal)

rc |> filter(traded, instrument == "BOND") |> arrange(spread_bp) |>
  transmute(ISIN = isin, Maturity = format(maturity_date, "%b %Y"),
            `Years` = round(ttm_years, 2), `Coupon (%)` = coupon_pct,
            `Market yield (%)` = round(yield_pct, 2),
            `Curve yield (%)` = round(model_yield, 2),
            `Difference (bp)` = spread_bp, Signal = signal) |>
  head(10) |>
  tbl(paste("Traded bonds ranked richest to cheapest against the fitted curve,",
            format(last_date, "%d %B %Y"), ". A positive difference means the bond yields more than the curve implies, so it is cheap."))
```

```{r fig-rich-cheap, fig.cap="Each security's market yield minus the yield implied by the fitted curve. Points above zero are cheap relative to the curve, points below are rich. Dashed lines mark plus and minus 50 basis points. Grey points did not trade, so a wide gap there usually means a stale mark rather than an opportunity."}
ggplot(rc, aes(ttm_years, spread_bp,
               colour = ifelse(traded, "Traded", "Not traded"))) +
  geom_hline(yintercept = 0, colour = red, linewidth = .7) +
  geom_hline(yintercept = c(-50, 50), colour = "grey80", linetype = 2) +
  geom_point(size = 2, alpha = .85) +
  scale_colour_manual(values = c(Traded = navy, `Not traded` = grey)) +
  labs(title = "Rich/cheap screen against the fitted curve",
       subtitle = paste("Market yield minus curve yield,", format(last_date, "%d %B %Y")),
       x = "Years to maturity", y = "Difference (basis points)",
       colour = NULL, caption = SRC)
```

## What a rate move costs a portfolio {#risk}

```{r risk}
duration_dv01 <- function(ytm, settle, maturity, coupon, freq = 2) {
  d <- coupon_dates(settle, maturity, freq)
  t <- yearfrac(settle, d); n <- t * freq
  cf <- rep(coupon / freq, length(d)); cf[length(cf)] <- cf[length(cf)] + 100
  y <- ytm / 100 / freq; pv <- cf / (1 + y)^n; dirty <- sum(pv)
  mac <- sum(t * pv) / dirty; mod <- mac / (1 + y)
  c(dirty = dirty, macaulay = mac, modified = mod, dv01 = mod * dirty / 10000)
}

port <- rc |> filter(instrument == "BOND", traded) |> mutate(face = 1e6 / n())
risk <- t(vapply(seq_len(nrow(port)), function(i)
  duration_dv01(port$yield_pct[i], last_date, port$maturity_date[i],
                port$coupon_pct[i]), numeric(4)))
port <- port |> mutate(modified_duration = risk[, "modified"],
                       dv01_per_100 = risk[, "dv01"],
                       dv01_ghs = dv01_per_100 * face / 100)

port |> transmute(ISIN = isin, `Years` = round(ttm_years, 2),
                  `Coupon (%)` = coupon_pct, `Yield (%)` = round(yield_pct, 2),
                  `Modified duration` = round(modified_duration, 2),
                  `Face (GHS)` = round(face),
                  `DV01 (GHS)` = round(dv01_ghs, 2)) |>
  tbl("An equally weighted GHS 1 million portfolio of the bonds that traded on the last day. DV01 is the loss from a one basis point rise in yields.")
```

```{r tab-scenarios}
shocks <- list(
  "Parallel +100bp"              = function(t) rep(100, length(t)),
  "Parallel -100bp"              = function(t) rep(-100, length(t)),
  "Steepener (1y -50, 10y +50)"  = function(t) approx(c(1, 10), c(-50, 50), t, rule = 2)$y,
  "Flattener (1y +50, 10y -50)"  = function(t) approx(c(1, 10), c(50, -50), t, rule = 2)$y,
  "Front-end selloff (+200bp under 2y)" = function(t) ifelse(t <= 2, 200, 0))

imap_dfr(shocks, ~ tibble(Scenario = .y,
                          `P&L (GHS)` = -sum(port$dv01_ghs * .x(port$ttm_years)))) |>
  mutate(`Share of face value` = percent(`P&L (GHS)` / 1e6, accuracy = 0.01)) |>
  tbl(paste0("Profit and loss on the GHS 1 million portfolio under five rate scenarios. Portfolio DV01 is GHS ",
             round(sum(port$dv01_ghs), 2), " per basis point."))
```

Table \@ref(tab:tab-scenarios) translates the PCA result into money. A general
rise in rates of one percentage point costs about twenty times what a
same-sized change in the curve's *shape* costs. For a treasury or pension desk,
that is the argument for hedging the level exposure first and worrying about
slope second.

# What this means, in plain terms {#implications}

```{r tab-implications}
tibble(
  `Who` = c("Government / Ministry of Finance", "Banks and pension funds",
            "Businesses borrowing", "Households and savers", "The market itself"),
  `What the results imply` = c(
    "Borrowing costs have fallen sharply across every maturity, and the curve now slopes up normally. Longer issuance has become viable again, though it is still priced at a premium to short borrowing.",
    "Interest rate risk is dominated by the general level of rates, not the curve's shape. A portfolio of government bonds gains or loses mostly when all rates move together, so level hedging matters most.",
    "Loan pricing benchmarks anchored to Treasury rates have fallen by double digits since 2023, which lowers the hurdle rate for investment.",
    "Treasury bills lost purchasing power through 2023 because inflation exceeded the interest earned. Since 2025 they have paid a clear positive real return.",
    "Securities of similar maturity are now priced far more consistently than in 2023. That is a sign of returning liquidity and confidence rather than a modelling artefact.")
) |>
  tbl("Implications of the results for different users of the yield curve.") |>
  wide()
```

# Limitations {#limitations}

```{r tab-limitations}
tibble(
  Limitation = c("Thin bond trading", "Long end is extrapolated",
                 "Pre-DDEP bonds excluded", "2023 curves are indicative",
                 "Inflation is monthly and lagged", "No credit or liquidity decomposition"),
  `What it means for the results` = c(
    "Roughly six in ten quote rows are carried-over marks. Weighting reduces their influence but cannot remove it.",
    paste0("The longest traded bond matures in about ", round(max(fits$max_ttm, na.rm = TRUE), 1),
           " years, so fitted rates beyond that are model extrapolation, not observation."),
    "The fitted curve describes the post-DDEP market. A robustness check including the old bonds would shift the long end up.",
    paste0("Median fit error in 2023 was about ",
           round(median(fits$rmse_bp[year(fits$report_date) == 2023])),
           " basis points, so levels from that year should be quoted with a margin."),
    "Real rates are computed ex post against published inflation, not against expected inflation, which is what a lender actually prices.",
    "The curve mixes expectations, term premium, credit and liquidity. Separating them would need instruments Ghana does not yet have, such as an index-linked bond.")
) |>
  tbl("Limitations of the analysis.") |>
  wide()
```

# Conclusion {#conclusion}

```{r tab-summary}
tibble(
  Objective = c("1. Assemble the data", "2. Establish conventions",
                "3. Fit daily curves", "4. Describe the evolution",
                "5. Apply the curve"),
  Result = c(
    paste0(format(nrow(quotes), big.mark = ","), " security-day quotes over ",
           n_distinct(quotes$report_date), " trading days, joined to auctions, the policy rate and inflation."),
    "Bills: money-market yield, actual/364, reproduces published yields exactly. Bonds: semi-annual, actual/365, quoted clean, median pricing error 0.03 per 100 on traded quotes.",
    paste0(nrow(fits), " curves fitted; Nelson-Siegel chosen over Svensson and bootstrapping for smoothness with comparable accuracy. Fit error fell from ",
           round(median(fits$rmse_bp[year(fits$report_date) == 2023])), "bp in 2023 to ",
           round(median(fits$rmse_bp[year(fits$report_date) == 2026])), "bp in 2026."),
    paste0("Inverted and distressed in 2023, flat in 2024, upward sloping by 2026. Level explains ",
           percent(explained[1], accuracy = 0.1), " of weekly changes; three factors explain ",
           percent(sum(explained[1:3]), accuracy = 0.1), ". The one-year real rate moved from deeply negative to clearly positive."),
    paste0("Rich/cheap screen on ", nrow(rc), " securities and a scenario analysis showing a GHS 1m bond portfolio with DV01 of GHS ",
           round(sum(port$dv01_ghs), 2), " per basis point, dominated by level risk."))
) |>
  tbl("Objectives and what the analysis delivered against each.") |>
  wide()
```

Three and a half years after the default, the price of Ghanaian government debt
looks like the price of debt in a working market again. The curve slopes upward,
savers earn a positive real return, and securities of similar maturity are
quoted at similar yields rather than scattered. The curve is not yet complete:
trading beyond ten years is thin, and the long end is as much model as market.
But the direction, and the speed, are unambiguous in the data.

The natural extensions are a term premium decomposition once a longer history
exists, a liquidity-adjusted curve that uses trade volume directly, and a
rolling out-of-sample test of whether the rich/cheap signal predicts
subsequent returns.

# Appendix: reproducibility {#appendix}

The report reads one workbook, `Ghana_Yield_Curve_Dataset_2023_2026.xlsx`,
produced by the data pipeline described in Section \@ref(data). Set
`fit_frequency: "daily"` in the YAML header to fit all
`r length(all_dates)` trading days instead of weekly.

```{r export, eval=FALSE}
write.csv(fits, "ghana_zero_curves.csv", row.names = FALSE)
write.csv(rc, paste0("rich_cheap_", last_date, ".csv"), row.names = FALSE)
```

```{r session}
sessionInfo()
```
